---
title: "s5e5_calorie_prediction"
author: "WY"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Target
The goal of this competition is to predict how many calories were burned during a workout.

ML tools: 
  tidymodels related worksflows & glm, lightgbm, ranger engine.
  
Evaluatioin metric:
  Root Mean Squared Error (RMSLE)
  
  sample_submission.csv : 
  basic lm score on rcp_base_v0 :
  kaggle best score is 0.0563 in the begining.

```{r}
library(yardstick)

# 核心计算逻辑（隐藏metric_impl参数）
rmsle_impl <- function(truth, estimate,case_weight=NULL) {
  sqrt(mean((log1p(truth) - log1p(estimate))^2))
}

# 向量化版本（处理NA值）
rmsle_vec <- function(truth, estimate, na_rm = TRUE,case_weights=NULL, ...) {
  check_numeric_metric(truth, estimate, case_weights)
  if (na_rm) {
    result <- yardstick_remove_missing(truth, estimate, case_weights)

    truth <- result$truth
    estimate <- result$estimate
    case_weights <- result$case_weights
  } else if (yardstick_any_missing(truth, estimate, case_weights)) {
    return(NA_real_)
  }
  rmsle_impl(truth = truth,
             estimate = estimate,
             case_weight=case_weigths)
}

# 数据框方法（支持分组计算）
rmsle <- function(data, ...) {
  UseMethod("rmsle")
}
rmsle <- new_numeric_metric(rmsle, direction = "minimize")

rmsle.data.frame <- function(data, truth, estimate, na_rm = TRUE,case_weights=NULL, ...) {
  numeric_metric_summarizer(
    name='rmsle',
    fn = rmsle_vec,
    data = data,
    truth = !!enquo(truth),
    estimate = !!enquo(estimate),
    na_rm = na_rm,
    case_weights=!!enquo(case_weights)
  )
}

rmsle_metric <- metric_set(rmse,rsq,rmsle)

```




  
Notice:


## Init library and Data

### library

```{r}
library(tidyverse)
library(tidymodels)
library(finetune)
library(future)
library(purrr)
library(furrr)
library(textrecipes)
library(themis)
library(embed)
#library(tailor)
library(ggfortify)


library(bonsai)
library(lightgbm)
library(xgboost)
library(ranger)
library(betareg)
library(readr)
library(janitor)
library(lubridate)
library(betareg)

library(memoise)
dk_cach <- memoise::cache_filesystem('./cache')
```
### download to local
```{r}
competition_name <- 'playground-series-s5e5'
data_path <- file.path('../input',competition_name)
target_name <- 'calories'
# system(paste0('kaggle competitions  download -c ', competition_name))
# unzip(paste0(competition_name,'.zip'),exdir=file.path('../input',competition_name))
# file.remove(paste0(competition_name,'.zip'))
```


### loading data

```{r}

train<- 
  readr::read_csv(file.path(data_path, 'train.csv'),
                  show_col_types = FALSE)|>
  janitor::clean_names()
test <- 
  readr::read_csv(file.path(data_path, 'test.csv'),
                  show_col_types = FALSE)|>
   janitor::clean_names()
submission <-  readr::read_csv(file.path(data_path, 'sample_submission.csv'),show_col_types = FALSE)
```

### quick skim

```{r}
#train|> skimr::skim()
```
important notice: 
- no missing data
- some column is not normal distribution

## EDA

## Data Preparing
### data split
```{r}

set.seed(42)
splits <- rsample::initial_split(train, prop = 0.8, strata = calories)
tr <- rsample::training(splits)
te <- rsample::testing(splits)
ncv  <- vfold_cv(tr,v=3,strata = calories)
```

### recipes

#### basic recipes (rmsle 0.5)
```{r}
rec_base <-  recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::check_missing(recipes::all_predictors())
```
#### v1_recipes (rmsle 0.37)
```{r}
rec_v1 <-  
  recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_mutate(
    bmi = weight / (height/100)^2, 
    age_squared = age^2,
    duration_squared = duration^2,
    heart_rate_squared = heart_rate^2,
    body_temp_squared = body_temp^2
  ) |>
  #recipes::step_interact(terms = ~ age:duration + height:weight) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::step_impute_knn(bmi, neighbors = 5) |>
  recipes::check_missing(recipes::all_predictors())
```

Explanation:

*   `bmi = weight / (height/100)^2`: Creates a Body Mass Index feature, which can capture the relationship between weight and height.  It's often more informative than the raw height and weight values.
*   `age_squared = age^2`: Adds a squared age term to capture potential non-linear relationships between age and calories burned.  The effect of age might not be linear, and this allows the model to fit a curve.
*   `duration_squared = duration^2`: Similar to age\_squared, this adds a squared duration term to capture non-linear effects of workout duration.
*   `step_interact(terms = ~ age:duration + height:weight)`: Includes interaction terms between age and duration, and height and weight. These interactions can capture the combined effect of these variables on calorie burn. For example, the impact of duration might be different for different age groups.
*   `step_impute_knn(bmi, neighbors = 5)`: Imputes missing BMI values using K-Nearest Neighbors imputation, using 5 neighbors.  This handles potential missing values in the created `bmi` feature.  It is placed before the `check_missing` step.

#### rec_v2 (rmsle need 0.155)
```{r}
rec_v2 <-   recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_mutate(
    bmi = weight / (height/100)^2, 
    age_squared = age^2,
    duration_squared = duration^2,
    heart_rate_squared = heart_rate^2,
    body_temp_squared = body_temp^2
  ) |>
  recipes::step_interact(terms = ~ age:duration + height:weight + heart_rate:body_temp) |>
  recipes::step_ns(age, deg_free = 3) |>
  recipes::step_ns(duration, deg_free = 3) |>
  recipes::step_ns(heart_rate, deg_free = 3) |>
  recipes::step_ns(body_temp, deg_free = 3) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::step_impute_knn(bmi, neighbors = 5) |>
  recipes::step_corr(threshold = 0.7) |>
  recipes::check_missing(recipes::all_predictors())
```

#### rec_v3(rmsle need better)
```{r}
rec_v3 <- recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_mutate(
    bmi = weight / (height/100)^2,
    age_squared = age^2,
    duration_squared = duration^2,
    heart_rate_squared = heart_rate^2,
    body_temp_squared = body_temp^2,
    age_bmi = age * bmi,
    duration_heart_rate = duration * heart_rate
  ) |>
  recipes::step_interact(terms = ~ age:duration + height:weight + heart_rate:body_temp ) |>
  recipes::step_ns(age, deg_free = 3) |>
  recipes::step_ns(duration, deg_free = 3) |>
  recipes::step_ns(heart_rate, deg_free = 3) |>
  recipes::step_ns(body_temp, deg_free = 3) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  #recipes::step_impute_knn(bmi, neighbors = 5) |>
  recipes::step_corr(threshold = 0.8) |>
  #recipes::step_pca(recipes::all_numeric_predictors(), threshold = 0.8) |>
  recipes::step_filter(calories > 0) |>
  #recipes::step_log(calories, base = 10) |>
  recipes::check_missing(recipes::all_predictors())
```

## quick model 
### linear regression

```{r}
current_rec <- rec_v2

lm_eng <- 
  parsnip::linear_reg(mode = "regression") |>
  parsnip::set_engine("lm")
wf_lm <- workflows::workflow() |>
workflows::add_recipe(current_rec) |>
workflows::add_model(lm_eng)



fit_lm <- wf_lm |> fit(tr)

aug_preds <- augment(fit_lm, te)|>mutate(.pred=case_when(.pred<=0~0,
                                                         .default=.pred))

aug_preds |> rmsle(truth=calories,estimate=.pred)|>print()
aug_preds|>select(.pred,calories) |> pivot_longer(cols=everything()) |>
  ggplot(aes(x=value, fill=name)) +
  geom_density(alpha=0.5)
```
```{r}
resample_option <- control_resamples(verbose=TRUE,save_workflow = TRUE, save_pred = TRUE,)
cv_wf <- wf_lm |>
  fit_resamples(ncv,
                metrics = rmsle_metric,
                control=resample_option
                      )#|>
#  collect_metrics()

```
### glm regression
```{r}


# 1. 定义Gamma回归模型
gamma_model <- linear_reg(
  penalty = 0.1,  # 无正则化
  mixture = 0.5   # 纯Gamma回归（需通过set_engine调整）
) %>%
  set_engine("glm", family = Gamma(link = "log"))  # 使用log链接函数

# 2. 创建工作流
wf <- workflow() %>%
  add_model(gamma_model) %>%
  add_recipe(current_rec)

# 3. 拟合模型
gamma_fit <- wf|> fit(data = tr)

# 4. 验证预测值是否为正

aug_preds <- augment(gamma_fit , te)|>mutate(.pred=case_when(.pred<=0~0,
                                                         .default=.pred))

aug_preds |> rmsle(truth=calories,estimate=.pred)|>print()
aug_preds|>select(.pred,calories) |> pivot_longer(cols=everything()) |>
  ggplot(aes(x=value, fill=name)) +
  geom_density(alpha=0.5)
```


### lightgbm regression

```{r}
current_rec <- rec_v2

lgbm_eng<-
   parsnip::boost_tree(
     #  trees = 300, # Number of trees
     #  learn_rate = 0.5,
     # # tree_depth =5,
     #  loss_reduction = 0.1,
     #  stop_iter = 50,
     #  #sample_size = 0.9, # Added sample_size
     #  #tree_depth = tune(),
     #  #mtry = 0.5,
     #  min_n = 100
   ) |>
   set_mode("regression")|>
   set_engine("lightgbm",
              #metric='roc_auc', 
              #num_leaves = 20,
              #counts = FALSE,
              #num_threads = 12,
              metric = "rmse",              # 优化目标
              # reg_alpha=0.01,
              # reg_lambda = 0.5,
              verbose=1) 

wf_lgb <- workflows::workflow() |>
workflows::add_recipe(current_rec) |>
workflows::add_model(lgbm_eng)


fit_lgb <- wf_lgb |> fit(tr)

aug_preds <- augment(fit_lgb, te)|>mutate(.pred=case_when(.pred<=0~0,
                                                         .default=.pred))

aug_preds |> rmsle(calories,.pred)|>print()
aug_preds|>select(.pred,calories) |> pivot_longer(cols=everything()) |>
  ggplot(aes(x=value, fill=name)) +
  geom_density(alpha=0.5)
```
### gam regression

```{r}
library(mgcv)
current_rec <- rec_v1 

tmp_df <- current_rec  |> prep() |>bake(new_data = tr) 
# rec <- recipe(mpg~displacement + horsepower +weight + acceleration + year, data =auto)|>
#   step_YeoJohnson(all_numeric_predictors())
```


```{r}
base_formula <- 
  calories ~ 

  s(age,k=10) +
  s(height,k=10) +
  s(weight,k=10) +
  s(duration,k=10)  +
  s(heart_rate,k=10) +
  s(body_temp,k=10) +
  s(bmi,k=10) +
  #te(bmi, age, k=c(5,5))+
  #te(height, weight,k=c(5,5)) +
  # te(age, height,k = c(5,5))+
  # te(age, weight,k = c(5,5))+
  # te(weight, height,k = c(5,5))+
  #te(duration, heart_rate,k = c(5,5))+
  #te(duration, weight,k=c(5,5))+
  #s(age_squared,k=10) +
  #s(heart_rate_squared,k=10) + 
 # s(body_temp_squared,k=10) +
 # s(duration_squared,k=10) + 
  #te(age, bmi,k=c(10,10))+
  #te(duration,heart_rate,k=c(10,10))+
  #te(heart_rate, body_temp,k=c(10,10)) 
  #s(age_x_duration) + s(height_x_weight) +
  #te(age, duration,k=c(10,10)) +
  factor(sex_male) 

v1_formula <-   calories ~  
  s(age) +
  #s(height,k=20) +
  #s(weight,k=20) +
  s(duration)  +
  s(heart_rate,k=10) +
  s(body_temp,k=10) +
  s(bmi,k=20) +
  factor(sex_male)  

v2_formula <-
  v1_formula |> 
  update(.~. + ti(duration, heart_rate) +
           ti(duration, heart_rate) +
           ti(duration, body_temp) +
           ti(duration, bmi)
  )
           
  


gam_formula <- v2_formula

bam_spec <-   bam(
  formula = gam_formula,  # 对交互项使用张量积te()
  data = tmp_df,
  response = TRUE, # 强制结果为正数。
  family = Gamma(link = "log"),
  select=TRUE,
  method = "fREML",    # 快速REML估计，速度比REML快3-10倍[4,6](@ref)
  discrete = TRUE,     # 启用离散化技术，降低内存占用
  nthreads = 12,        # 建议设置为物理核心数（非超线程数）
  chunk.size = 10**5 # 分块处理数据，减少内存峰值
)
bam_spec|> summary()
bam_spec |>glance()

# pred_gam <- predict(gam_wf, new_data = tr)
aug_preds <- augment(bam_spec, tmp_df,type.predict="response")|>rename(.pred= .fitted)#|>
 # mutate(.pred=case_when(.pred<=0~0, .default=.pred))

aug_preds |> rmsle(calories,.pred)|>print()
aug_preds|>select(.pred,calories) |> pivot_longer(cols=everything()) |>
  ggplot(aes(x=value, fill=name)) +
  geom_density(alpha=0.5)


```



#### purrr compare
```{r}
library(furrr)
library(mgcv)
current_rec <- rec_v1 

tmp_df <- current_rec  |> prep() |>bake(new_data = tr) 
get_result<- function(gam_form){
  df <- tmp_df# |>head(100000)
  bam_spec <-   bam(
    formula = gam_form,  # 对交互项使用张量积te()
    data = df,
    response = TRUE, # 强制结果为正数。
    family = Gamma(link = "log"),
    select=TRUE,
    method = "fREML",    # 快速REML估计，速度比REML快3-10倍[4,6](@ref)
    discrete = TRUE,     # 启用离散化技术，降低内存占用
    nthreads = 12,        # 建议设置为物理核心数（非超线程数）
    chunk.size = 10**5 # 分块处理数据，减少内存峰值
  ) 
  #aug_preds <- augment(bam_spec, df,type.predict="response")|>rename(.pred= .fitted)
  #rmsle_value <- aug_preds |> rmsle(calories,.pred)|>pull('.estimate',1)
  #return_value <- bam_spec|> glance() |> mutate(rmsle=rmsle_value)
  return(bam_spec)
}

get_rmsle <- function(model, df) {
  tmp_df <-df 
  aug_preds <- augment(model, tmp_df,type.predict="response")|>rename(.pred= .fitted)
  result <- aug_preds |> rmsle(calories,.pred)
  return(result)
}



form_base=calories ~ s(age,k=15) +s(duration,k=25) + s(heart_rate,k=30) + s(body_temp,k=15,bs='cr') + s(bmi, bs = "ps", k = 50, m = c(2, 2)) + factor(sex_male)

form_dur_heart_weight= form_base|> update(.~. + te(duration, heart_rate,k = c(5,5)) +  te(duration, weight,k=c(5,5)))

v3_formula <-   calories ~s(age) +  + s(duration)  + s(heart_rate,k=10) + s(body_temp,k=10) + s(bmi,k=20) + factor(sex_male)  

v4_formula <- v3_formula |> update(.~. + ti(duration, heart_rate) + ti(duration, heart_rate) + ti(duration, body_temp) +       ti(duration, bmi)   )
v9_formula <- calories ~
  s(age, k = 10) + # OK
  s(duration, k = 20) + # OK
  s(heart_rate, k = 25) + # Was k=20, edf=18.69, p=0.010. Increase.
  s(body_temp, k = 15) + # OK
  s(bmi, k = 30) + # Was k=25, edf=20.03, p<2e-16. Increase.

  factor(sex_male) + # Or your actual dummy variable name

  # ti(duration, heart_rate, weight, by = age):
  # Was k=c(6,6,6) -> k'=(6-1)^3 = 125, edf=120.07, p<2e-16.
  # Try k=c(7,7,7) -> k'=(7-1)^3 = 216.
  ti(duration, heart_rate, weight, by = age, k = c(7, 7, 7)) +

  ti(duration, body_temp, k = c(7, 7)) + # OK with k'=(7-1)^2 = 36

  # ti(duration, bmi):
  # Was k=c(7,7) -> k'=(7-1)^2 = 36, edf=33.71, p=0.045.
  # Try k=c(8,8) -> k'=(8-1)^2 = 49.
  ti(duration, bmi, k = c(8, 8)) +

  # ti(bmi, duration, heart_rate):
  # Was k=c(6,6,6) -> k'=(6-1)^3 = 125, edf=122.62, p=0.005.
  # Try k=c(7,7,7) -> k'=(7-1)^3 = 216.
  ti(bmi, duration, heart_rate, k = c(7, 7, 7))

v11b_formula <-  form_base|> update(.~.+   ti(duration, heart_rate, weight, by = age, k = c(9, 9, 9)) +
                                      ti(duration, body_temp, k = c(7, 7)) +
                                      ti(duration, bmi, k = c(12, 12)) +
                                      ti(bmi, duration, heart_rate, k = c(9, 9, 9)))

v12_formula <- form_base |>update(.~. + ti(duration, heart_rate, weight, by = age, bs = c("ps", "ps", "ps"), k = c(7, 7, 7)) + 
                                    ti(duration, body_temp, k = c(7,7), bs = c("cr", "cr")) + 
                                    ti(duration, bmi, bs = c("ps", "ps"), k = c(8, 8)) +
                                      # ti(bmi, duration, heart_rate): Similar logic.
                                    ti(bmi, duration, heart_rate, bs = c("ps", "ps", "ps"), k = c(7, 7, 7)))
                                    
param_list<- list(base=form_base, # 0.9833
                  #v1=form_dur_heart_weight # bad module  0.9612. worse than base
                  #v3 = v3_formula,
                  #v4= v4_formula
                  v11b_formula =v11b_formula,
                  v12 = v12_formula
                  )
mod_list <-
  param_list |>
  map(\(x) get_result(x))
mod_list|>saveRDS(file='mod_list.rds')
mod_list|>map_dfr(\(x) glance(x),.id='name')

```

##### diagnose
```{r}
col_names<- tmp_df |>names()
mod_list <- readRDS(file='mod_list.rds')
plot_df <- mod_list$v4 |>
  augment(newdata = tmp_df,type.predict="response") |> 
  rename(.pred= .fitted) |>
  mutate(.residual=log1p(calories)-log1p(.pred)) 

for (col in col_names) {
  plot_df |>ggplot(aes(x=.data[[col]], y=.residual)) +geom_smooth() |> print()
  
}
#after heart_rate >125 has high residuals *** very important it might be sharp change. need consider the sline mode. also heart_rate_squared has similar impact and heavier.
## heart_rate improved a bit at v4 model

# bmi at right side, when value increase the .residual change bigger.
#height, weight has high residual 


```
## kaggle submission
Prepare Submission ----

```{r}
set.seed(1234)

#library(future)
#plan(multisession,workers = 12)
#final_model <- simple_wf_fit|>extract_workflow()

#final_model <- wf_lm |>fit(tr)#|>extract_workflow()
final_model <- bam_spec 
final_predictions <- final_model |>
   predict(newdata = current_rec|>prep()|>bake(test),type="response") #|>
   #mutate(.pred=case_when(.pred<0~abs(.pred),
    #                      .default=.pred))
#plan(sequential)

 # # #Handle negative predictions
 #  final_predictions <- final_predictions |>
 #    mutate(.pred= ifelse(.pred< 0, 0, abs(.pred)))

 # Save submission file
 submission |>
   bind_cols(.pred=final_predictions)|>
   mutate(Calories =.pred) |>
   select(-.pred)|>
   readr::write_csv("submission.csv")
 zip('submission.csv.zip','submission.csv')
 
```



### score submit
```{r}
# # submit latest submission.csv
# system('kaggle competitions submit -c playground-series-s5e5 -f submission.csv.zip -m "0.0914"')
# 
# Sys.sleep(15)
# # get latest score 
# system('kaggle competitions submissions -q -c playground-series-s5e5')

```

### notebook convert
```{r}
 # library(rmd2jupyter)
 # rmd2jupyter('calories_predict.Rmd')
```


