---
title: "s5e5_calorie_prediction"
author: "WY"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Target
The goal of this competition is to predict how many calories were burned during a workout.

ML tools: 
  tidymodels related worksflows & glm, lightgbm, ranger engine.
  
Evaluatioin metric:
  Root Mean Squared Error (RMSLE)
  
  sample_submission.csv : 
  basic lm score on rcp_base_v0 :
  kaggle best score is 0.0563 in the begining.

```{r}
library(yardstick)

# 核心计算逻辑（隐藏metric_impl参数）
rmsle_impl <- function(truth, estimate,case_weight=NULL) {
  sqrt(mean((log1p(truth) - log1p(estimate))^2))
}

# 向量化版本（处理NA值）
rmsle_vec <- function(truth, estimate, na_rm = TRUE,case_weights=NULL, ...) {
  check_numeric_metric(truth, estimate, case_weights)
  if (na_rm) {
    result <- yardstick_remove_missing(truth, estimate, case_weights)

    truth <- result$truth
    estimate <- result$estimate
    case_weights <- result$case_weights
  } else if (yardstick_any_missing(truth, estimate, case_weights)) {
    return(NA_real_)
  }
  rmsle_impl(truth = truth,
             estimate = estimate,
             case_weight=case_weigths)
}

# 数据框方法（支持分组计算）
rmsle <- function(data, ...) {
  UseMethod("rmsle")
}
rmsle <- new_numeric_metric(rmsle, direction = "minimize")

rmsle.data.frame <- function(data, truth, estimate, na_rm = TRUE,case_weights=NULL, ...) {
  numeric_metric_summarizer(
    name='rmsle',
    fn = rmsle_vec,
    data = data,
    truth = !!enquo(truth),
    estimate = !!enquo(estimate),
    na_rm = na_rm,
    case_weights=!!enquo(case_weights)
  )
}

rmsle_metric <- metric_set(rmse,rsq,rmsle)

```




  
Notice:


## Init library and Data

### library

```{r}
library(tidyverse)
library(tidymodels)
library(finetune)
library(future)
library(purrr)
library(furrr)
library(textrecipes)
library(themis)
library(embed)
#library(tailor)
library(ggfortify)


library(bonsai)
library(lightgbm)
library(xgboost)
library(ranger)
library(betareg)
library(readr)
library(janitor)
library(lubridate)
library(betareg)

library(memoise)
dk_cach <- memoise::cache_filesystem('./cache')
```
### download to local
```{r}
competition_name <- 'playground-series-s5e5'
data_path <- file.path('../input',competition_name)
target_name <- 'calories'
# system(paste0('kaggle competitions  download -c ', competition_name))
# unzip(paste0(competition_name,'.zip'),exdir=file.path('../input',competition_name))
# file.remove(paste0(competition_name,'.zip'))
```


### loading data

```{r}

train<- 
  readr::read_csv(file.path(data_path, 'train.csv'),
                  show_col_types = FALSE)|>
  janitor::clean_names()
test <- 
  readr::read_csv(file.path(data_path, 'test.csv'),
                  show_col_types = FALSE)|>
   janitor::clean_names()
submission <-  readr::read_csv(file.path(data_path, 'sample_submission.csv'),show_col_types = FALSE)
```

### quick skim

```{r}
#train|> skimr::skim()
```
important notice: 
- no missing data
- some column is not normal distribution

## EDA

## Data Preparing
### data split
```{r}

set.seed(42)
splits <- rsample::initial_split(train, prop = 0.8, strata = calories)
tr <- rsample::training(splits)
te <- rsample::testing(splits)
ncv  <- vfold_cv(tr,v=3,strata = calories)
```

### recipes

#### basic recipes (rmsle 0.5)
```{r}
rec_base <-  recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::check_missing(recipes::all_predictors())
```
#### v1_recipes (rmsle 0.37)
```{r}
rec_v1 <-  
  recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_mutate(
    bmi = weight / (height/100)^2, 
    age_squared = age^2,
    duration_squared = duration^2
  ) |>
  recipes::step_interact(terms = ~ age:duration + height:weight) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::step_impute_knn(bmi, neighbors = 5) |>
  recipes::check_missing(recipes::all_predictors())
```

Explanation:

*   `bmi = weight / (height/100)^2`: Creates a Body Mass Index feature, which can capture the relationship between weight and height.  It's often more informative than the raw height and weight values.
*   `age_squared = age^2`: Adds a squared age term to capture potential non-linear relationships between age and calories burned.  The effect of age might not be linear, and this allows the model to fit a curve.
*   `duration_squared = duration^2`: Similar to age\_squared, this adds a squared duration term to capture non-linear effects of workout duration.
*   `step_interact(terms = ~ age:duration + height:weight)`: Includes interaction terms between age and duration, and height and weight. These interactions can capture the combined effect of these variables on calorie burn. For example, the impact of duration might be different for different age groups.
*   `step_impute_knn(bmi, neighbors = 5)`: Imputes missing BMI values using K-Nearest Neighbors imputation, using 5 neighbors.  This handles potential missing values in the created `bmi` feature.  It is placed before the `check_missing` step.

#### rec_v2 (rmsle need 0.155)
```{r}
rec_v2 <-   recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_mutate(
    bmi = weight / (height/100)^2, 
    age_squared = age^2,
    duration_squared = duration^2,
    heart_rate_squared = heart_rate^2,
    body_temp_squared = body_temp^2
  ) |>
  recipes::step_interact(terms = ~ age:duration + height:weight + heart_rate:body_temp) |>
  recipes::step_ns(age, deg_free = 3) |>
  recipes::step_ns(duration, deg_free = 3) |>
  recipes::step_ns(heart_rate, deg_free = 3) |>
  recipes::step_ns(body_temp, deg_free = 3) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::step_impute_knn(bmi, neighbors = 5) |>
  recipes::step_corr(threshold = 0.7) |>
  recipes::check_missing(recipes::all_predictors())
```

#### rec_v3(rmsle need better)
```{r}
rec_v3 <- recipes::recipe(calories ~ ., data = tr) |>
  recipes::step_rm(id) |>
  recipes::step_mutate(
    bmi = weight / (height/100)^2,
    age_squared = age^2,
    duration_squared = duration^2,
    heart_rate_squared = heart_rate^2,
    body_temp_squared = body_temp^2,
    age_bmi = age * bmi,
    duration_heart_rate = duration * heart_rate
  ) |>
  recipes::step_interact(terms = ~ age:duration + height:weight + heart_rate:body_temp ) |>
  recipes::step_ns(age, deg_free = 3) |>
  recipes::step_ns(duration, deg_free = 3) |>
  recipes::step_ns(heart_rate, deg_free = 3) |>
  recipes::step_ns(body_temp, deg_free = 3) |>
  recipes::step_dummy(recipes::all_nominal_predictors()) |>
  recipes::step_zv(recipes::all_predictors()) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  #recipes::step_impute_knn(bmi, neighbors = 5) |>
  recipes::step_corr(threshold = 0.8) |>
  #recipes::step_pca(recipes::all_numeric_predictors(), threshold = 0.8) |>
  recipes::step_filter(calories > 0) |>
  #recipes::step_log(calories, base = 10) |>
  recipes::check_missing(recipes::all_predictors())
```

## quick model 
### linear regression



```{r}
current_rec <- rec_v2
lm_eng <- 
  parsnip::linear_reg(mode = "regression") |>
  parsnip::set_engine("lm")
wf_lm <- workflows::workflow() |>
workflows::add_recipe(current_rec) |>
workflows::add_model(lm_eng)


fit_lm <- wf_lm |> fit(tr)

aug_preds <- augment(fit_lm, te)|>mutate(.pred=case_when(.pred<=0~0,
                                                         .default=.pred))

aug_preds |> rmsle(truth=calories,estimate=.pred)|>print()
aug_preds|>select(.pred,calories) |> pivot_longer(cols=everything()) |>
  ggplot(aes(x=value, fill=name)) +
  geom_density(alpha=0.5)
```
```{r}
resample_option <- control_resamples(verbose=TRUE,save_workflow = TRUE, save_pred = TRUE,)
cv_wf <- wf_lm |>
  fit_resamples(ncv,
                metrics = rmsle_metric,
                control=resample_option
                      )#|>
#  collect_metrics()

```

### lightgbm regression

```{r}
current_rec <- rec_v2

lgbm_eng<-
   parsnip::boost_tree(
     #  trees = 300, # Number of trees
     #  learn_rate = 0.5,
     # # tree_depth =5,
     #  loss_reduction = 0.1,
     #  stop_iter = 50,
     #  #sample_size = 0.9, # Added sample_size
     #  #tree_depth = tune(),
     #  #mtry = 0.5,
     #  min_n = 100
   ) |>
   set_mode("regression")|>
   set_engine("lightgbm",
              #metric='roc_auc', 
              #num_leaves = 20,
              #counts = FALSE,
              #num_threads = 12,
              metric = "rmse",              # 优化目标
              # reg_alpha=0.01,
              # reg_lambda = 0.5,
              verbose=1) 

wf_lgb <- workflows::workflow() |>
workflows::add_recipe(current_rec) |>
workflows::add_model(lgbm_eng)


fit_lgb <- wf_lgb |> fit(tr)

aug_preds <- augment(fit_lgb, te)|>mutate(.pred=case_when(.pred<=0~0,
                                                         .default=.pred))

aug_preds |> rmsle(calories,.pred)|>print()
aug_preds|>select(.pred,calories) |> pivot_longer(cols=everything()) |>
  ggplot(aes(x=value, fill=name)) +
  geom_density(alpha=0.5)
```
### gam regression

```{r}
library(tidymodels)
library(mgcv)
gam_spec <- 
  gen_additive_mod(
    select_features = FALSE,  # 禁用自动特征选择
    adjust_deg_free = tune()  # 允许自由度调优
  ) %>%
  set_engine(
    "mgcv", 
    method = "fREML",         # 使用快速REML优化方法
    discrete = TRUE,          # 启用离散方法加速计算
    nthreads = 8,            # 并行线程数
    family = gaussian()      # 指定分布族（根据任务调整）
  ) %>%
  set_mode("regression")     # 或 "classification"
gam_wf <- workflow() %>%
  add_recipe(rec_v1) %>%
  add_model(gam_spec)
gam_formula <- calories ~ 
  s(duration, bs = "cr", k = 15) +          # 三次样条捕捉运动时长非线性效应[6,8](@ref)
  s(heart_rate, bs = "ad", k = 10) +       # 自适应平滑处理心率异方差性[4](@ref)
  ti(age, duration, bs = "cr") +           # 年龄与运动时长的张量交互平滑[6](@ref)
  s(bmi, bs = "tp") +                      # 薄板样条拟合BMI复杂关系[6](@ref)
  s(body_temp, bs = "cr", k = 8) +         # 体温的局部平滑
  sex_male +                               # 性别线性效应
  s(age_x_duration) +                      # 交互项平滑扩展[6](@ref)
  s(height_x_weight)    
# 心率与体温交互项平滑
# 使用bam加速计算（59万+样本）


gam_wf_fit <- gam_wf |>fit(formula=gam_formula, data=tr)
```

```{r}
library(mgcv)
current_rec <- rec_v1 
tmp_df <- current_rec  |> prep() |>bake(new_data = tr) 

gam_model <- bam(
  gam_formula,
  data = tmp_df,                    # 使用recipes预处理后的数据
  family = Gamma(link = "log"),            # 卡路里为正值，Gamma分布更优[4](@ref)
  method = "REML",                         # 自动平滑参数选择[8](@ref)
  discrete = TRUE,                         # 离散化加速
  nthreads = 8,                            # 并行计算
  gamma = 1.5                              # 增强正则化防过拟合[8](@ref)
)

pred_gam <- predict(gam_model, newdata = rec_v1|>prep()|>bake(te), type = "response")

# 交叉验证
#cv_results <- cv.gam(gam_model, nfolds=5, parallel=TRUE)


```

```{r}
pred_gam <- predict(gam_model, newdata =current_rec|>prep()|>bake(tr), type = "response")
aug_preds <- tr|>mutate(.pred_gam = pred_gam)
```


```{r}
# 绘制残差诊断图
#par(mfrow = c(2,2))
#gam.check(gam_model, rep = 500)            # 检查残差分布与QQ图[6](@ref)
```

```{r}
cv_splits <- vfold_cv(juice(rec_v1|>prep()), v = 5)
cv_results <- fit_resamples(
  gam_model,
  resamples = cv_splits,
  metrics = rmsle_metric,
  option=resample_option
)
```

## kaggle submission
Prepare Submission ----

```{r}
set.seed(1234)

#library(future)
#plan(multisession,workers = 12)
#final_model <- simple_wf_fit|>extract_workflow()

final_model <- wf_lm |>fit(tr)#|>extract_workflow()
final_predictions <- final_model |>
   predict(new_data = test) #|>
   #mutate(.pred=case_when(.pred<0~abs(.pred),
    #                      .default=.pred))
#plan(sequential)

 # #Handle negative predictions
  final_predictions <- final_predictions |>
    mutate(.pred= ifelse(.pred< 0, 0, abs(.pred)))

 # Save submission file
 submission |>
   bind_cols(final_predictions)|>
   mutate(Calories =.pred) |>
   select(-.pred)|>
   readr::write_csv("submission.csv")
 zip('submission.csv.zip','submission.csv')
 
```



### score submit
```{r}
# submit latest submission.csv
system('kaggle competitions submit -c playground-series-s5e5 -f submission.csv.zip -m "0.155"')

Sys.sleep(15)
# get latest score 
system('kaggle competitions submissions -q -c playground-series-s5e5')

```

### notebook convert
```{r}
 library(rmd2jupyter)
 rmd2jupyter('podcast_listening_time.Rmd')
```


